---
title: "Data Visualization"
description: |
 This R script creates several plots to visualize the data. Mixed effects regression models are also applied.
author:
  - first_name: "Ayush"
    last_name: "Noori"
    url: https://www.github.com/ayushnoori
    affiliation: Massachusetts General Hospital
    affiliation_url: https://www.serranopozolab.org
    orcid_id: 0000-0003-1420-1236
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages and define directories.

```{r load-packages, message=FALSE, warning=FALSE}

# data manipulation
library(data.table)
library(purrr)
library(magrittr)

# data visualization
library(ggplot2)
library(ggpubr)

# mixed effects model
library(lmerTest)
library(openxlsx)

# utility functions
library(brainstorm)

```

Note that directories are relative to the R project path.

```{r define-directores}

# set directories
dir4 = file.path("Results", "4 - Spectral Clustering")
dir6 = file.path("Results", "6 - Data Visualization")

```

# Load Data

Load processed ROI measurement data from the `4 - Spectral Clustering` directory.

```{r load-data}

all = readRDS(file.path(dir4, "Z-Score Data.rds"))

```

# Define Plotting Functions

Define variables which contain theme options.

```{r set-themes}

# define theme options
marker_theme = theme(
  plot.title = element_text(size = 16, hjust = 0.5, face = "bold.italic"),
  
  # axes
  axis.title.y = element_text(size = 12, face = "bold"),
  axis.title.x = element_blank(),
  axis.text.x = element_text(size = 12, face = "bold", color = "black"),
  axis.ticks.x = element_blank(),
  
  # panel
  panel.background = element_rect(fill = "#EBEBEB"),
  panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
  panel.grid = element_blank(),
  
  # legend
  legend.title = element_text(size = 14, face = "bold"),
  legend.text = element_text(size = 12),
  legend.position = "bottom")


# modify theme for proportions
prop_theme = marker_theme + theme(
  legend.position = "right",
  plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
  legend.text = element_text(size = 10, face = "plain"),
  panel.grid.major.y = element_line(size = 0.4, color = "#333333"),
  strip.text = element_text(size = 12, face = "bold", color = "white"),
  strip.background = element_rect(color = "black", fill = "#3D3D3D",
                                  size = 0.8, linetype = "solid"))

# modify theme for bar graph
bg_theme = prop_theme + theme(
  panel.grid.major.y = element_line(size = 0.2, color = "#333333"),
  panel.grid.minor.y = element_line(size = 0.2, color = "#333333"))


# modify theme for histogram
hist_theme = prop_theme + theme(
  axis.text.x = element_text(size = 10, face = "plain", color = "black"),
  axis.text.y = element_text(size = 10, face = "plain", color = "black"),
  axis.ticks.x = element_line(size = 0.5, color = "black"),
  axis.ticks.y = element_line(size = 0.5, color = "black"))


# modify theme for density
density_theme = hist_theme

```

Define plotting functions to visualize data by marker, state, and layer.

```{r boxplot-functions}

# create boxplots for each marker
plot_marker = function(dat, mx, grp, grpcol, legend = FALSE,
                       fname = NULL, facet = NULL) {
  
  # base plot
  marker_plot = ggplot(dat, aes(x = get(grp), y = get(mx), color = get(grp))) +
    geom_boxplot() +
    scale_color_manual(grp, values = levels(dat[[grpcol]])) + 
    ggtitle(mx) +
    labs(x = grp, y = "Mean Gray Intensity (MGI)")
  
  # split by layer and assign appropriate theme
  if(!is.null(facet)) {
    marker_plot = marker_plot +
      facet_wrap(. ~ Layer, nrow = 1,
                 labeller = function(x) return(map(x, ~paste("Layer", .x)))) +
      prop_theme + theme(plot.title = element_text(size = 16, hjust = 0.5,
                                                   face = "bold.italic"))
  } else {
    marker_plot = marker_plot + marker_theme
  }
  
  # plot legend
  if(!legend) { marker_plot = marker_plot + theme(legend.position = "none") }
  
  # save file
  if(!is.null(fname)) { ggsave(paste0(fname, ".pdf"), marker_plot, width = 4, height = 6) }
  
  # return modified plot
  return(marker_plot + theme(axis.text.x = element_blank(), plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")))
  
}

# shared y-axis
shared_y = function(p, idx, nwidth) {if((idx - 1) %% nwidth == 0) return(p) else return(p + rremove("ylab"))}

# arrange plots
arrange_plots = function(plist, leg, lab, grplab, nwidth = ceiling(length(plist)/2), nheight = ceiling(length(plist)/nwidth), title = TRUE) {
  
  # create shared y-axis per row
  plist = imap(plist, ~shared_y(.x, .y, nwidth))
  
  # join plots together
  composite = ggarrange(plotlist = plist, ncol = nwidth, nrow = nheight, legend.grob = leg, legend = "bottom")
  
  # add title
  if(title) { composite = annotate_figure(composite, top = text_grob(paste(lab, "Marker Expression by", grplab), size = 20, face = "bold")) }
  
  # return composite
  return(composite)
  
}

```

Plot stacked boxplots where ROIs are split by Control/Alzheimer, then grouped by stratified Distance, Layer, or some other grouping variable. Relative proportions within each group are visualized.

```{r proportion-functions}

plot_prop = function(dat, grpvar, grpcol, lab, fname = NULL) {
  
  grpcols = c(grpvar, "Condition", "State", grpcol)
  
  # group by grouping variable
  prop = dat %>%
    .[, .N, by = grpcols] %>%
    .[order(.[, ..grpcols]), ] %>%
    .[, Proportion := map(.(N), ~.x*100/sum(N)), by = c(grpvar, "Condition")] %>%
    .[, Label := paste0(round(Proportion, 1), "%")]
  
  if(grpvar == "Layer") { prop[, Layer := paste("Layer", Layer)] }
  
  # plot proportions data
  prop_plot = ggplot(prop, aes(x = get(grpvar), y = Proportion,
                               fill = State, label = Label)) +
    geom_bar(position = "stack", stat = "identity", width = 0.5,
             color = "black", size = 0.4) +
    facet_grid(. ~ Condition) +
    scale_fill_manual(grpvar, values = levels(prop[[grpcol]])) +
    scale_y_continuous(expand = expansion(mult = c(0, 0))) +
    geom_text(size = 3, position = position_stack(vjust = 0.5)) +
    # ggtitle(paste(lab, "State Across", grpvar)) +
    labs(x = gsub("Bin", "", grpvar), y = "Proportion", fill = "State") +
    prop_theme
  
  if(!is.null(fname)) ggsave(paste0(fname, ".pdf"), prop_plot, width = 14, height = 6) else return(prop_plot)
  
}

```

Plot bargraph where ROIs are split by Control/Alzheimer, then grouped by stratified Distance, Layer, or some other grouping variable. Relative proportions within each phenotypic state are visualized (or, phenotypic state within the grouping variable; i.e., x-axis can be phenotypic state or other grouping variable).

```{r bargraph-functions}

plot_bg = function(dat, grpvar, grpcol, lab, fname = NULL, xvar = "State") {
  
  grpcols = c("Condition", xvar, grpvar, grpcol)
  
  # group by grouping variable
  bg = dat %>%
    .[, .N, by = grpcols] %>%
    .[order(.[, ..grpcols]), ] %>%
    .[, Proportion := map(.(N), ~.x*100/sum(N)), by = c("Condition")] %>%
    .[, Label := paste0(round(Proportion, 1), "%")] %>%
    .[, Fraction :=  map(.(N), ~paste0(.x, "/", sum(N))), by = c("Condition")]
  
  # plot proportions data
  bg_plot = ggplot(bg, aes(x = get(xvar), y = Proportion,
                           fill = get(grpvar), label = Fraction)) +
    geom_bar(stat = "identity", position = position_dodge(0.6),
             width = 0.4, color = "black", size = 0.4) +
    facet_grid(. ~ Condition) +
    scale_fill_manual(grpvar, values = levels(bg[[grpcol]])) + 
    scale_y_continuous(expand = expansion(mult = c(0, .1))) +
    geom_text(size = 2.5, position = position_dodge(0.6), vjust = -1.5) +
    # ggtitle(paste(lab, grpvar, "Across State")) +
    labs(x = gsub("Bin", "", xvar), y = "Proportion") +
    bg_theme
  
  if(!is.null(fname)) ggsave(paste0(fname, ".pdf"), bg_plot, width = 14, height = 6) else return(bg_plot)
  
}

```

Define function to plot histogram of distance.

```{r histogram-functions}

# plot grouped density
plot_density = function(dat, grpvar, grpcol, linecol, lab, maxlim) {
  
  density_plot = ggplot(dat, aes(x = get(grpvar), group = State)) +
    geom_density(aes(fill = State), alpha = 0.5) +
    scale_fill_manual(values = levels(dat[[grpcol]])) +
    geom_density(color = "black", size = 0.3) +
    # geom_density(aes(color = State)) +
    # scale_color_manual(values = levels(dat[[linecol]])) +
    labs(x = grpvar, y = "Density") +
    scale_x_continuous(expand = c(0, 0), breaks = seq(0, 500, by = 50)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    density_theme + theme(legend.position = "none")
  
  return(density_plot)
  
}


# plot individual histogram
plot_hist = function(dat, grpvar, grpcol, lab, stateidx = 1:3, legend = F,
                     bin_width = 20, t_d = NULL, max_x = 500, max_y = 1) {
  
  hist_plot = ggplot(dat, aes(x = get(grpvar), fill = State)) +
    geom_histogram(
      aes(y = ..count../t_d),
      # aes(y = ..density..),
      breaks = seq(0, max(dat[, ..grpvar] + bin_width), by = bin_width),
      position = "identity", size = 0.3,
      color = "black") +
    scale_fill_manual(values = levels(dat[[grpcol]])[stateidx]) +
    # geom_density(fill = "white", alpha = 0.2) +
    labs(x = grpvar, y = "Proportion", fill = "State") +
    scale_x_continuous(limits = c(0, max_x), expand = c(0.02, 0.02)) +
    scale_y_continuous(limits = c(0, max_y), 
                       expand = expansion(mult = c(0, 0.05))) +
    facet_wrap(. ~ State, scales = "free_y") +
    hist_theme + theme(legend.position = "bottom")
  
  if(legend) return(hist_plot) else return(hist_plot +
                                             theme(legend.position = "none"))
  
}


# get maximum y-value
max_y = function(dat, bin_width, t_d, total_denominator = F) {
  
  if(total_denominator) t_d = t_d[State == dat[1, State], N]
  
  dat = dat[, !c("State")]
  
  maximum_y = dat %>%
    { cut(.[[1]], breaks = seq(0, max(. + bin_width), by = bin_width),
          include.lowest = T) } %>%
    summary() %>% { ./t_d } %>% max() %>% return()
  
}


# plot composite histograms
plot_hists = function(my_dat, grpvar, grpcol, lab, fname = NULL,
                      condvar = c("Control", "Alzheimer"),
                      smplvar = NULL,
                      bin_width = 25, total_denominator = F) {
  
  if(!is.null(smplvar)) my_dat = my_dat[Sample %in% smplvar]
  
  # should the denominator be ALL astrocytes within Condition?
  og_total_denominator = nrow(my_dat[Condition %in% condvar])
  
  # should the denominator be TOTAL Homeostatic astrocytes for Homeostatic plot,
  # TOTAL Intermediate astrocytes for Intermeidate plot, etc. (within Condition)
  og_state_denominator = my_dat[Condition %in% condvar, .N, by = "State"]
  
  # REMOVE crops WITHOUT plaques (but keep whole denominator)
  dat = copy(my_dat) %>%
    .[is.na(get(grpvar)), c(grpvar) := 500] %>%
    .[get(grpvar) > 500, c(grpvar) := 500] %>%
    .[Condition %in% condvar]
  
  # create density plot
  density_plot = plot_density(dat, grpvar, grpcol, "StateColors",
                              lab, maxlim)
  
  # get maximum x-value
  max_x = max(dat[, ..grpvar], na.rm = T)
  
  # create individual histograms
  if(total_denominator) {
    
    # get maximum y-value
    max_y = dat[, max_y(.SD, bin_width, og_total_denominator, F),
                  .SDcols = c(grpvar, "State"), by = "State"][, max(V1)]
    
    # plot histogram
    hist_plots = imap(levels(dat[, State]),
                      ~plot_hist(dat[State == .x], grpvar, grpcol, lab,
                                 stateidx = .y, bin_width = bin_width,
                                 t_d = og_total_denominator,
                                 max_x = max_x, max_y = max_y))
  
  } else  { 
    
    # get maximum value
    max_y = dat[, max_y(.SD, bin_width, og_state_denominator, T),
                  .SDcols = c(grpvar, "State"), by = "State"][, max(V1)]
    
    # plot histogram
    hist_plots = imap(levels(dat[, State]),
                      ~plot_hist(dat[State == .x], grpvar, grpcol, lab,
                                 stateidx = .y, bin_width = bin_width,
                                 t_d = og_state_denominator[State == .x, N],
                                 max_x = max_x, max_y = max_y))
  
  }
  
  # get composite legend
  hist_legend = get_legend(plot_hist(dat, grpvar, grpcol, lab, legend = T, t_d = nrow(dat)))
  
  # create composite histogram
  comp_plot = arrange_plots(rev(hist_plots), hist_legend, nwidth = 3, title = F)
  
  # add density plot
  comp_plot = ggarrange(density_plot, comp_plot, ncol = 1)
  
  # save or return plot
  if(!is.null(fname)) ggsave(paste0(fname, ".pdf"), comp_plot, width = 10, height = 8) else return(comp_plot)
  
}

```

# Mixed Effects Models

Function to run mixed effects regression models.

```{r mixed-effects}

mixed_effects = function(dat, my_formula, model_name,
                         ldir = NULL, marker = "N/A") {
  
  # run mixed model
  my_model = lmerTest::lmer(my_formula, REML = T, data = dat)
  my_summary = summary(my_model)
  my_confint = confint(my_model)
  
  # get model results
  mcols = c("Marker", "Model", "Comparison", "Estimate", "Pr(>|t|)", "Std. Error")
  
  # model results
  model_results = my_summary$coefficients %>%
    as.data.table(keep.rownames = "Comparison") %>%
    .[, Marker := ..marker] %>%
    .[, Model := ..model_name] %>%
    .[, .SD, .SDcols = mcols] %>%
    merge(as.data.table(my_confint, keep.rownames = "Comparison"),
          by = "Comparison")
  
  # write output
  if(!is.null(ldir)) {
    sink(file = ldir, append = T)
    cat("\n", rep("_", 80), sep = "")
    cat("\n\n\n>>>", toupper(model_name), "MODEL:\n\n"); print(my_model)
    cat("\n\n>>> SUMMARY:\n\n"); print(my_summary)
    cat("\n\n>>> CONFIDENCE INTERVALS:\n\n"); print(my_confint)
    sink()
  }
  
  # change names and set order
  setnames(model_results, c("2.5 %", "97.5 %"), c("Lower CI", "Upper CI"))
  setcolorder(model_results, c("Marker", "Model"))
  return(model_results)
  
}

# run multiple models per marker
mixed_marker = function(marker, dat, mxdir) {
  
  # create file
  ldir = file.path(mxdir, paste(marker, "Mixed Effects Model.txt"))
  if(file.exists(ldir)) file.remove(ldir) else file.create(ldir)
  
  # log message
  sink(file = ldir, append = T)
  cat(">>> MARKER: ", marker, "\n", sep = "")
  sink()
  
  # run model 1
  m1 = mixed_effects(dat, get(marker) ~ State + (1|Sample), "State", ldir, marker)
  m2 = mixed_effects(dat, get(marker) ~ Condition + (1|Sample), "Condition",
                     ldir, marker)
  
  # final model results
  mres = rbind(m1, m2)
  
  return(mres)
  
}

```

# Define Iteration Functions

Define function to create plots.

```{r plot-data}

plot_data = function(dat, lab, mx, pcols, pwb) {
  
  # create subdirectory if needed
  wdir = file.path(dir6, lab)
  mxdir = file.path(wdir, "Mixed Effects Models")
  hdir = file.path(wdir, "Distance Histograms")
  bgdir = file.path(wdir, "Distance Bar Graphs")
  if(!dir.exists(wdir)) {dir.create(wdir); dir.create(mxdir);
    dir.create(hdir); dir.create(bgdir)}
  
  # group None with > 50
  hmdists = colnames(dat) %>% .[grep("Bin", .)]
  group_none = function(x) dat[get(x) == "None", c(x) := "> 50 um"]
  walk(hmdists, group_none)
  
  # define plotting colors
  dat = dat %>%
    .[, StateColors := factor(State, labels = pcols$State)] %>%
    .[, StatePastelColors := factor(State, labels = pcols$StatePastel)] %>%
    .[, SampleColors := factor(Sample, labels = pcols$Sample)] %>%
    .[, ConditionColors := factor(Condition, labels = pcols$Condition)] %>%
    .[, ConditionPastelColors := factor(Condition,
                                        labels = pcols$ConditionPastel)] %>%
    .[, LayerColors := factor(Layer, labels = pcols$Layer)] %>%
    .[, DistanceBinColors := factor(DistanceBin, labels = pcols$DistanceBin)]
  
  
  ## MIXED EFFECTS MODELS
  
  # run models
  mxef = map(mx, ~mixed_marker(.x, dat, mxdir))
  mxef = rbindlist(mxef)
  
  # add distance mixed models
  m3 = mixed_effects(dat, Distance ~ State + (1|Sample), "Distance")
  m4 = mixed_effects(dat[Condition == "Control"],
                     Distance ~ State + (1|Sample), "Distance in CTRL")
  m5 = mixed_effects(dat[Condition == "Alzheimer"],
                     Distance ~ State + (1|Sample), "Distance in AD")
  
  # bind mixed model data
  mxef = rbind(mxef, m3, m4, m5)
  
  # write to file
  brainstorm::add_worksheet(pwb, lab, mxef)
  wb_colors = list(# State = "#DCE5DF", Condition = "#E4DEDD",
    State = "#BDDBC8", Condition = "#E2D0CA",
    Distance = "#E0EAF5", `Distance in CTRL` = "#D1E0F0", `Distance in AD` = "#C1D6EB")
  iwalk(wb_colors, ~addStyle(pwb, lab, createStyle(fgFill = .x),
                             rows = which(mxef$Model == .y) + 1,
                             cols = 2, stack = T))
  
  ## SUMMARY PLOT
  cond_dat = copy(dat) %>%
    .[, .N, by = c("State", "Condition")] %>%
    .[order(.[, c("State", "Condition")]), ] %>%
    .[, Proportion := map(.(N), ~.x*100/sum(N)), by = c("Condition")] %>%
    .[, Label := paste0(round(Proportion, 1), "%")]
  
  state_dat = copy(dat) %>%
    .[, .N, by = c("State", "Condition")] %>%
    .[order(.[, c("State", "Condition")]), ] %>%
    .[, Proportion := map(.(N), ~.x*100/sum(N)), by = c("State")] %>%
    .[, Label := paste0(round(Proportion, 1), "%")]
  
  
  # plot proportions data
  cond_plot = ggplot(cond_dat, aes(x = Condition, y = Proportion, fill = State, label = Label)) +
    geom_bar(position = "stack", stat = "identity", ,
             width = 0.5, color = "black", size = 0.4) +
    scale_fill_manual(values = levels(dat[["StatePastelColors"]])) + 
    scale_y_continuous(expand = expansion(mult = c(0, 0))) +
    geom_text(size = 3, position = position_stack(vjust = 0.5)) +
    labs(x = "Condition", y = "Proportion") +
    bg_theme + theme(legend.position = "none")
  
  # plot proportions data
  state_plot = ggplot(state_dat, aes(x = State, y = Proportion, fill = Condition, label = Label)) +
    geom_bar(position = "stack", stat = "identity", ,
             width = 0.5, color = "black", size = 0.4) +
    scale_fill_manual(values = levels(dat[["ConditionPastelColors"]])) + 
    scale_y_continuous(expand = expansion(mult = c(0, 0))) +
    geom_text(size = 3, position = position_stack(vjust = 0.5)) +
    labs(x = "State", y = "Proportion") +
    bg_theme + theme(legend.position = "none")
  
  comb_plot = ggarrange(cond_plot, state_plot, ncol = 1)
  
   ggsave(file.path(wdir, "Proportions Summary.pdf"), comb_plot, width = 4, height = 12)
  
  
  ## CONDITION/STATE BOXPLOTS
  
  # create boxplots
  condition_plots = map(mx, ~plot_marker(dat, .x, "Condition", "ConditionColors"))
  state_plots = map(mx, ~plot_marker(dat, .x, "State", "StateColors"))
  
  # get legends
  condition_legend = get_legend(plot_marker(dat, mx[1], "Condition", "ConditionColors", legend = TRUE))
  state_legend = get_legend(plot_marker(dat, mx[1], "State", "StateColors", legend = TRUE))
  
  # create composite plots
  ggsave(file.path(wdir, "Marker Expression by Condition.pdf"), arrange_plots(condition_plots, condition_legend, lab, "Condition"),
         width = ceiling(length(mx)/2)*3, height = 10)
  
  ggsave(file.path(wdir, "Marker Expression by State.pdf"), arrange_plots(state_plots, state_legend, lab, "State"),
         width = ceiling(length(mx)/2)*3, height = 10)
  
  
  ## LAYER BOXPLOTS
  
  # create boxplots
  layer_condition_plots = map(mx, ~plot_marker(dat, .x, "Condition", "ConditionColors", facet = "Layer"))
  layer_state_plots = map(mx, ~plot_marker(dat, .x, "State", "StateColors", facet = "Layer"))
  
  # get legends
  layer_condition_legend = get_legend(plot_marker(dat, mx[1], "Condition", "ConditionColors", legend = TRUE, facet = "Layer"))
  layer_state_legend = get_legend(plot_marker(dat, mx[1], "State", "StateColors", legend = TRUE, facet = "Layer"))
  
  # create composite plots
  ggsave(file.path(wdir, "Marker Expression by Condition per Layer.pdf"),
         arrange_plots(layer_condition_plots, layer_condition_legend, lab,
                       "Condition per Layer", nwidth = 2),
         width = 24, height = length(mx)*3)
  
  ggsave(file.path(wdir, "Marker Expression by State per Layer.pdf"),
         arrange_plots(layer_state_plots, layer_state_legend, lab,
                       "State per Layer", nwidth = 2),
         width = 24, height = length(mx)*3)

  
  ## BAR GRAPHS
  
  # define distance metrics
  dists = c("Distance")
  
  # plot bar graphs for state
  walk(dists, ~plot_bg(dat, "State", "StatePastelColors", lab, file.path(bgdir, paste("State Within", .x)), paste0(.x, "Bin")))
  
  # plot bar graphs for distance
  walk(dists, ~plot_bg(dat, paste0(.x, "Bin"), "DistanceBinColors", lab, file.path(bgdir, paste(.x, "Within State")), "State"))
  
  
  ## HISTOGRAMS
  
  walk(dists, ~plot_hists(dat, .x, "StatePastelColors", lab, fname = file.path(hdir, paste(.x, "in Select CTRL with Total Denominator")), condvar = c("Control"), smplvar = c("2169", "2250"), bin_width = 25, total_denominator = T))
  
  walk(dists, ~plot_hists(dat, .x, "StatePastelColors", lab, fname = file.path(hdir, paste(.x, "in CTRL with Total Denominator")), condvar = c("Control"), bin_width = 25, total_denominator = T))
  
      walk(dists, ~plot_hists(dat, .x, "StatePastelColors", lab, fname = file.path(hdir, paste(.x, "in AD with Total Denominator")), condvar = c("Alzheimer"), bin_width = 25, total_denominator = T))
    
  # return data
  return(dat)
  
}

```

# Create Plots

Create the plots specified in `plot_data` by mapping over `all`.

```{r create-plots}

# define markers of interest
markers = list(
  Astrocyte = c("GFAP", "YKL40", "VIM", "TSPO", "EAAT1", "EAAT2", "GS"),
  Microglia = c("MHC2", "CD68", "TMEM119", "TSPO", "FTL"),
  Vessel = c("GFAP", "YKL40", "VIM", "TSPO", "EAAT1", "EAAT2", "GS"))

# define distance colors
distcols = c('< 25 um' = "#FFAF85", '25-50 um' = "#FFED85", '> 50 um' = "#96BDD9")

# define color palette
cols = list(
  DistanceBin = distcols,
  Layer = c(II = "#DDF2B2", III = "#8DD2B9", IV = "#39AEC3", V = "#2072B1", VI = "#0C2C84"),
  Sample = c('1190' = "#A6CEE3", '1301' = "#5D9FC9", '1619' = "#2A7FB0", '1684' = "#79B79A", '1820' = "#9ED57B", '2124' = "#5AB348", '2148' = "#619E45", '2157' = "#CC9B7F", '2169' = "#F37272", '2191' = "#E62D2F", '2207' = "#ED593B", '2242' = "#FBB268", '2250' = "#FDA13B", '2274' = "#FF7F00"),
  Condition = c(Control = "#377EB8", Alzheimer = "#CE6D8B"),
  State = c('Homeostatic' = "#39B200", 'Intermediate' = "#F0C808", 'Reactive' = "#960200"),
  StatePastel = c('Homeostatic' = "#9ECC7F", 'Intermediate' = "#EADA86", 'Reactive' = "#B67977"),
  ConditionPastel = c(Control = "#B0C6DE", Alzheimer = "#E4C2CC")
)

# create plots
wb = createWorkbook()
plots = imap(all, ~plot_data(.x, .y, markers[[.y]], cols, wb))
saveWorkbook(wb, file.path(dir6, "Mixed Effects Models.xlsx"), overwrite = TRUE)

```

# Distance to Plaques

Now, get summary statistics for distance to plaques to compute Chi-squared test.

```{r count-distance}

count_distance = function(dat, lab, pwb) {
  
  # create data frames
  close_plaques = dat[Distance <= 50, .N, by = c("Condition", "State")]
  far_plaques = dat[Distance > 50 | is.na(Distance), .N,
                    by = c("Condition", "State")]
  
  # set names
  setnames(close_plaques, "N", "Close to Plaques/Tangles")
  setnames(far_plaques, "N", "Far from Plaques/Tangles")
  
  # merge data
  pdata = merge(close_plaques, far_plaques, by = c("Condition", "State"), sort = F)
  
  # add to Excel
  brainstorm::add_worksheet(pwb, lab, pdata)
  
}

```

Get summary statistics by mapping over `all`.

```{r get-distance}

# create plots
plaque_wb = createWorkbook()
plots = imap(all, ~count_distance(.x, .y, plaque_wb))
saveWorkbook(plaque_wb, file.path(dir6, "Distance to Neuropathology Proportions.xlsx"), overwrite = TRUE)

```
