---
title: "Parse ROI Measurements"
description: |
 This R script parses the measurements obtained from the previous ROI segmentation script in ImageJ.
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages and define directories. Note that this script uses my personal utilities package `brainstorm`, which can be downloaded via `devtools::install_github("ayushnoori/brainstorm")`.

```{r load-packages, message=FALSE, warning=FALSE}

# data manipulation
library(data.table)
library(purrr)
library(magrittr)

# data visualization
library(ggplot2)

# Excel manipulation
library(openxlsx)

# utility functions
library(brainstorm)

```

Note that directories are relative to the R project path.

```{r define-directores}

# set directories
ddir = file.path("Data", "3 - ROIs")
dir3 = file.path("Results", "3 - ROI Measurements")
dir3.1 = file.path(dir3, "3.1 - Normalization Plots")

```

# Define Functions

Define function to retrieve the coordinate data, then convert from pixels to microns based on the crop resolution metadata.

```{r get-coordinates}

get_coordinates = function(cname, fname) {
  
  # read coordinate and resolution data
  coords = fread(file.path(fname, paste0(cname, "_ROIs.csv")))
  res = fread(file.path(fname, paste0(cname, "_Resolution.txt")))$V1
  
  # calculate center of VIA annotation relative to crop
  coords %>%
    .[, CenterX := X + Width/2] %>%
    .[, CenterY := Y + Height/2]
  
  # convert coordinates from pixels to microns (approx. 6.1 pixels : 1 micron)
  coords[, c("Width", "Height", "CenterX", "CenterY") := map(.(Width, Height, CenterX, CenterY), ~.x/res)]
  
  # replace old X and Y coordinates with new coordinates
  coords = coords[, .(Name, Width, Height, CenterX, CenterY,
                      Type, Quality, Annotator)]
  setnames(coords, c("CenterX", "CenterY"), c("X", "Y"))
  
  return(coords)
  
}

```

Define function to calculate the distance to the nearest plaque or tangle, where `coord` is the coordinate pair for a specific ROI, while `neuropath` is a `data.table` containing parsed plaque and/or tangle ROI data. Note that distances cannot be less than `0`. Also, the `Radius` for tangle ROIs is set to `0` in the subsequent chunk.

```{r compute-distance}

# function to compute distance
compute_distance = function(x, y, neuropath) {
  
  if(nrow(neuropath) == 0) return(NA) else {
    
    neuropath = neuropath %>% 
    .[, Raw := sqrt((X - x)^2 + (Y - y)^2)] %>%
    .[, Distance := Raw - Radius] %>%
    .[Distance < 0, Distance := 0]
  
  return(neuropath[, min(Distance)]) 
    
  }
  
}

# function to assign distance based on filter, which must be wrapped in expr()
assign_distance = function(listobj, lab, filter) { 
  
  # list contains both data and neuropathology objects
  dat = listobj[[1]]
  neuropath = listobj[[2]]
  
  # assign distance
  dat[, (lab) := pmap_dbl(dat[, .(X, Y)], ~compute_distance(.x, .y, neuropath[eval(filter), ]))]
  
  # return new list
  return(invisible(list(dat, neuropath)))
  
}

```

Define function to extract and aggregate ROI data from individual crops.

```{r parse-crop}

parse_crop = function(fname) {
  
  # extract crop attributes
  cinfo = strsplit(fname, "/")[[1]]
  cname = cinfo[4]; csplit = strsplit(cname, "_")[[1]]
  message(paste(c("-", csplit), collapse = " "))
  
  # read and parse data
  dat = fread(file.path(fname, paste0(cname, "_Measurements.csv"))) %>% 
    .[, ROI := map_chr(strsplit(Label, ":"), 1)] %>%
    .[, Marker := map_chr(strsplit(Label, ":"), 3)]
  
  # reshape data from long to wide to extract MGI
  wdat = dat[, .(ROI, Marker, Mean)] %>% dcast(ROI ~ ..., value.var = "Mean")
  
  # keep Area and Perimeter variables, overwrite long data
  dat = dat[!duplicated(ROI), .(ROI, Area, Perim.)] %>% 
    merge(wdat, ., by = "ROI", all.x = TRUE)
    
  # populate with metadata
  dat %>%
    .[, ID := paste(cname, ROI, sep="_")] %>%
    .[, Group := gsub("[0-9]", "", ROI)] %>%
    .[, Number := as.numeric(gsub("[a-zA-Z]", "", ROI))] %>%
    .[, Condition := cinfo[3]] %>%
    .[, Sample := csplit[1]] %>%
    .[, Layer := gsub("Layer", "", csplit[2])] %>%
    .[, Crop := gsub("crop", "", csplit[3])]
  
  # join coordinates with ROI measurements
  dat = get_coordinates(cname, fname) %>% 
    merge(dat, ., by.x = "ROI", by.y = "Name", all.x = TRUE)
  
  # separate neuropathology ROIs and calculate radius
  neuropath = dat %>%
    .[Group %in% c("Plaque", "Tangle"), ] %>%
    .[, .(Group, Type, ROI, Area, X, Y, Width, Height)] %>%
    .[, Radius := sqrt(Area/pi)] %>%
    .[Group == "Tangle", Radius := 0]
  
  # remove neuropathology
  dat = dat[!(Group %in% c("Plaque", "Tangle")), ]
  
  # compute distance
  list(dat, neuropath) %>%
    assign_distance("Distance", expr(Group %in% c("Plaque", "Tangle"))) %>%
    assign_distance("Plaque", expr(Group == "Plaque")) %>%
    assign_distance("Large", expr(Group == "Plaque" & Area > 50)) %>%
    assign_distance("Compact", expr(Group == "Plaque" & Type == "compact")) %>%
    assign_distance("Diffuse", expr(Group == "Plaque" & Type == "diffuse")) %>%
    assign_distance("Tangle", expr(Group == "Tangle")) %>%
    assign_distance("Intraneuronal", expr(Group == "Tangle" & Type == "intra")) %>%
    assign_distance("Extraneuronal", expr(Group == "Tangle" & Type == "extra"))
  
  # set column order
  setcolorder(dat, c("ROI", "ID", "Group", "Number", "Condition",
                     "Sample", "Layer", "Crop"))
  
  return(dat[, ROI := NULL])
  
}

```

# Parse ImageJ ROI Data

Map the `parse_crop` function over the list of crops measured by ImageJ.

```{r map-crops}

# get crop list
crops = list.files(file.path(ddir, c("CTRL", "AD")), full.names = TRUE)

# map over crop list
message("Parsing ROI Data:")
output = map_dfr(crops, ~parse_crop(.x))

# convert condition factor and order ROIs
output = output %>%
  .[, Condition := factor(Condition, levels = c("CTRL", "AD"), labels = c("Control", "Alzheimer"))] %>%
  .[order(Condition, Sample, Layer, Crop, Group, Number), ]

```

# Normalize Data

Rename certain columns to create syntactically valid names.

```{r clean-data}

# rename specific columns
setnames(output, c("Ferritin", "HuC/D", "PHF1-tau", "Vimentin", "Perim."), c("FTL", "HuC.D", "PHF1.tau", "VIM", "Perimeter"))

# get marker list
metadata = c("ID", "Group", "Number", "Condition", "Sample", "Layer", "Crop",
             "Area", "Perimeter", "Width", "Height", "X", "Y", "Type", "Quality",
             "Annotator", "Distance", "Plaque", "Large", "Compact",
             "Diffuse", "Tangle", "Intraneuronal", "Extraneuronal")
markers = colnames(output) %>% .[!(. %in% metadata)]

```

Normalize mean gray intensity (MGI) values by applying a `log`-transformation and computing z-scores.

```{r normalize-data}

# function to compute z-scores.
compute_z = function(x) { return((x-mean(x))/sd(x)) }

# copy non-normalized data
raw = copy(output)

# normalize data
output[, (markers) := map_dfc(.SD, ~compute_z(log(.x + 1))),
       .SDcols = markers, by = .(Group)]

# show output
show_table(output[, map(.SD, ~mean(.x)), .SDcols = markers, by = Group])
show_table(output[, map(.SD, ~sd(.x)), .SDcols = markers, by = Group])

```

# Save Data

Save output and display table. Tables in Excel are styled with the `openxlsx` package. Visualize data normalization by plotting histograms of raw and normalized MGI values.

```{r save-output}

fwrite(output, file.path(dir3, "ROI Measurements.csv"))

show_table(output[sample(nrow(output), 40), ])

```

Plot histograms of before and after normalization.

```{r plot-normalization}

plot_data = function(dat_long, lab) {
  
  p = ggplot(dat_long, aes(x = value, fill = Condition)) +
    geom_histogram(bins = 30, alpha = 0.5, color = "black") +
    facet_wrap(~ variable, ncol = 6, scales = "free") +
    scale_fill_manual(values = c("#377EB8", "#CE6D8B")) + 
    labs(title = lab,
         x = "Normalized Mean Gray Intensity",
         y = "Frequency",
         fill = "Condition") +
    theme(plot.title = element_text(hjust = 0.5, size = 16, face="bold"),
          axis.title.x = element_text(size=14, face="bold"),
          axis.title.y = element_text(size=14, face="bold"),
          legend.title = element_text(size=12, face="bold"),
          legend.text = element_text(size=10), legend.position = "bottom",
          strip.text = element_text(size=10, face="bold"),
          strip.background = element_rect(color="black", fill="#D9D9D9",
                                          size=1, linetype="solid"),
          panel.border = element_rect(color = "black", fill = NA, size = 1))
  
}

# plot raw data
raw_long = melt(raw, id.vars = c("ID", "Condition", "Group"),
                measure.vars = markers)
raw_plot = plot_data(raw_long, "Pre-Normalization Histograms")
print(raw_plot)
ggsave(file.path(dir3, "Pre-Normalization Histograms.pdf"),
       raw_plot, width = 24, height = 12)

# plot normalized data
output_long = melt(output, id.vars = c("ID", "Condition", "Group"),
                   measure.vars = markers)
output_plot = plot_data(output_long, "Post-Normalization Histograms")
print(output_plot)
ggsave(file.path(dir3, "Post-Normalization Histograms.pdf"),
       output_plot, width = 24, height = 12)

```

Save data to a formatted Excel file for readability.

```{r write-excel}

# create workbook
wb = createWorkbook()
sname = "ROI Measurements"

# header for metadata
hs1 = createStyle(fgFill = "#A37C40", fontColour = "#FFFFFF", fontName = "Arial Black", halign = "center", valign = "center", textDecoration = "Bold", border = "Bottom", borderStyle = "thick", fontSize = 14)

# header for markers
hs2 = createStyle(fgFill = "#1D3557", fontColour = "#FFFFFF", fontName = "Arial Black", halign = "center", valign = "center", textDecoration = "Bold", border = "Bottom", borderStyle = "thick", fontSize = 14)

# create worksheet
tcols = ncol(output)
addWorksheet(wb, sheetName = sname)
writeDataTable(wb, sname, x = output, tableStyle = "TableStyleMedium15",
               bandedRows = FALSE)
setColWidths(wb, sname, cols = 1:tcols, widths = "auto")
setColWidths(wb, sname, cols = 8:24, widths = 12)
setColWidths(wb, sname, cols = 25:26, widths = 14)
setColWidths(wb, sname, cols = 27:28, widths = 18)
setColWidths(wb, sname, cols = 34:tcols, widths = 22)
setRowHeights(wb, sname, rows = (1:nrow(output))+1, heights = 18)
freezePane(wb, sname, firstActiveRow = 2, firstActiveCol = 8)

# style headers
addStyle(wb, sname, hs1, rows = 1, cols = c(1:7, 25:tcols))
addStyle(wb, sname, hs2, rows = 1, cols = 8:24)

# style marker data
addStyle(wb, sname, createStyle(fontColour = "#1A1D23", fgFill = "#FFFFFF",
                                fontName = "Arial", fontSize = 10,
                                halign = "center", valign = "center"),
         rows = which(1:nrow(output) %% 2 == 0) + 1, cols = 8:24, gridExpand = TRUE)

addStyle(wb, sname, createStyle(fontColour = "#1A1D23", fgFill = "#F3F4F6",
                                fontName = "Arial", fontSize = 10,
                                halign = "center", valign = "center"),
         rows = which(1:nrow(output) %% 2 != 0) + 1, cols = 8:24, gridExpand = TRUE)

# style other columns
addStyle(wb, sname, createStyle(fontColour = "#1A1D23", fgFill = "#F6F4F4",
                                fontName = "Arial", fontSize = 10,
                                halign = "center", valign = "center"),
         rows = 1:nrow(output) + 1, cols = c(1:7, 25:tcols), gridExpand = TRUE)

# style metadata

# astrocyte metadata
addStyle(wb, sname, createStyle(fontColour = "#1A1D23", fgFill = "#FDEDEE",
                                fontName = "Arial", fontSize = 10,
                                halign = "center", valign = "center"),
         rows = which(output$Group == "Astrocyte") + 1, cols = c(1:7),
         gridExpand = TRUE)

# astrocyte header
addStyle(wb, sname, createStyle(fontColour = "#FFFFFF", fgFill = "#C98686",
                                fontName = "Arial", textDecoration = "Bold",
                                fontSize = 10, halign = "center",
                                valign = "center"),
         rows = which(output$Group == "Astrocyte") + 1, cols = 2,
         gridExpand = TRUE)

# microglia metadata
addStyle(wb, sname, createStyle(fontColour = "#1A1D23", fgFill = "#F4F6F4",
                                fontName = "Arial", fontSize = 10,
                                halign = "center", valign = "center"),
         rows = which(output$Group == "Microglia") + 1, cols = c(1:7),
         gridExpand = TRUE)

# microglia header
addStyle(wb, sname, createStyle(fontColour = "#FFFFFF", fgFill = "#708B75",
                                fontName = "Arial", textDecoration = "Bold",
                                fontSize = 10, halign = "center",
                                valign = "center"),
         rows = which(output$Group == "Microglia") + 1, cols = 2,
         gridExpand = TRUE)

# vessel metadata
addStyle(wb, sname, createStyle(fontColour = "#1A1D23", fgFill = "#F1F6F9",
                                fontName = "Arial", fontSize = 10,
                                halign = "center", valign = "center"),
         rows = which(output$Group == "Vessel") + 1, cols = c(1:7),
         gridExpand = TRUE)

# vessel header
addStyle(wb, sname, createStyle(fontColour = "#FFFFFF", fgFill = "#457B9D",
                                fontName = "Arial", textDecoration = "Bold",
                                fontSize = 10, halign = "center",
                                valign = "center"),
         rows = which(output$Group == "Vessel") + 1, cols = 2, gridExpand = TRUE)

# add conditional formatting
for(i in 34:tcols) {
  conditionalFormatting(wb, sname, cols = i, rows = 1:nrow(output) + 1,
                        type = "colourScale",
                        style = c("#D9A3A3", "#F8F6F6", "#8DAE93"))
}
  
# save workbook
saveWorkbook(wb, file.path(dir3, "ROI Measurements.xlsx"), overwrite = TRUE)

```
