---
title: "Plot ROC Curves"
description: |
 This R script plots the receiver operating characteristic (ROC) curves for the convolutional neural network models and calculates the area under the ROC curves (AUC).
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages and define directories.

```{r load-packages, message=FALSE, warning=FALSE}

# data manipulation
library(data.table)
library(purrr)
library(magrittr)

# data visualization
library(ggplot2)

# ROC curve
library(pROC)
library(plotROC)

# utilities
library(brainstorm)

```

Note that directories are relative to the R project path.

```{r define-directores}

# create file structure
celltypes = c("Astrocyte", "Microglia") %>% purrr::set_names()

# set directories
ddir = c("2 - Astrocyte CNN", "3 - Microglia CNN") %>% file.path("Results", "CNN", ., "Output") %>% purrr::set_names(celltypes)
dir4 = file.path("Results", "4 - Spectral Clustering")
dir8 = file.path("Results", "8 - CNN Interpretability")

```

# Load Data

Define function to read CNN output.

```{r read-cnn}

# function to read CNN output
read_cnn = function(fpath, lab) {
  
  # select most recent CNN output file
  fname = list.files(fpath, pattern = "\\.csv$") %>%
  .[strsplit(., "_") %>% { order(map_chr(., 2), map_chr(., 1)) }] %>% .[length(.)]
  
  # print and read file
  cat(paste0("\n", toupper(lab), "\nTarget Directory: ", fpath, "\nInput File: ", fname, "\n"))
  if (length(fname) > 0) return(fread(file.path(fpath, fname))) else return(NULL)

}

```

Load processed ROI measurement data from the `4 - Spectral Clustering` directory and CNN output from the `CNN\2 - Astrocyte CNN\Output` directory.

```{r load-data}

# read ROI data
all = readRDS(file.path(dir4, "Z-Score Data.rds"))[names(celltypes)]

# read CNN output
cnn = imap(ddir, read_cnn)

```

# Merge Data

Define function to merge ROI measurement data and clustering metadata with CNN output.

```{r merge-data}

# function to parse and merge data
merge_data = function(allx, cnnx) {
  
  # parse CNN data
  cnnx %>%
    .[, c("V1", "Image") := NULL] %>%
    .[, File := strsplit(File, "/")] %>%
    .[, File := map(File, ~tail(.x, 1))] %>%
    .[, ID := gsub("(\\.tif|AD_|CTRL_)", "", File)] %>%
    .[, PredictedLabel := factor(PredictedLabel, levels = c(1, 0), labels = c("Control", "Alzheimer"))] %>%
    .[, TrueLabel := factor(TrueLabel, levels = c(1, 0), labels = c("Control", "Alzheimer"))]
  
  # merge data
  setcolorder(cnnx, "ID")
  cnnx = merge(cnnx, allx, by = "ID", all.x = TRUE, all.y = FALSE)
  return(cnnx)
  
}

```

Map function over data objects for cell-types with CNN output data.

```{r map-merge}

# remove null CNN data
keep = names(which(!map_lgl(cnn, is.null)))
celltypes = celltypes[keep]; all = all[keep]; cnn = cnn[keep]

# function to parse and merge data
all = map(celltypes, ~merge_data(all[[.x]], cnn[[.x]]))

```

# Plot Data

Define function to plot histogram.

```{r plot-histogram}

# function to plot histogram
plot_histogram = function(dat, grp, grpcol, facet_grp = NULL, pos = "identity", nbins = 30, density = FALSE) {
  
  p = ggplot(dat, aes(x = ProbabilityAD, fill = get(grp))) +
    geom_histogram(position = pos, bins = nbins, alpha = 0.5, color = "black") +
    scale_fill_manual(values = levels(dat[[grpcol]])) + 
    labs(x = "Alzheimer Classification Probability", y = "Count", fill = grp) +
    theme(axis.title.x = element_text(size=14, face="bold"), axis.title.y = element_text(size=14, face="bold"),
          legend.title = element_text(size=12, face="bold"), legend.text = element_text(size=10), legend.position = "bottom",
          strip.text = element_text(size=16, face="bold"),
          strip.background = element_rect(color="black", fill="#D9D9D9", size=1, linetype="solid"),
          panel.border = element_rect(color = "black", fill = NA, size = 1))
  
  if(!is.null(facet_grp)) {
    p = p +
      facet_wrap(~ get(facet_grp), ncol = 3) +
      geom_density(aes(y=..density.. * 15), fill = "white", alpha = 0.3, linetype = "dashed")
  }
  
  if(density) {
    p = p + geom_density(aes(y=..density.. * 10, color = get(grp)), fill = "white", alpha = 0.3, linetype = "dashed") +
      scale_color_manual(values = levels(dat[[grpcol]])) + labs(color = grp)
  }
  
  return(p)
  
}

```

Define function to create plots.

```{r plot-data}

plot_data = function(dat, lab, pcols) {
  
  # create subdirectory if needed
  wdir = file.path(dir8, lab)
  if(!dir.exists(wdir)) {dir.create(wdir)}
  
  # define plotting colors
  dat = dat %>%
    .[, StateColors := factor(State, labels = pcols$State)] %>%
    .[, SampleColors := factor(Sample, labels = pcols$Sample)] %>%
    .[, ConditionColors := factor(Condition, labels = pcols$Condition)]
  
  # create histograms
  p_state = plot_histogram(dat, "State", "StateColors", "State")
  p_condition = plot_histogram(dat, "Condition", "ConditionColors", nbins = 50, density = FALSE)
  p_cs = plot_histogram(dat, "Condition", "ConditionColors", "State")
  
  # save histograms
  ggsave(file.path(wdir, "State Classification Probabilities.pdf"), p_state, width = 16, height = 6)
  ggsave(file.path(wdir, "Condition Classification Probabilities.pdf"), p_condition, width = 6, height = 6)
  ggsave(file.path(wdir, "Condition + State Classification Probabilities.pdf"), p_cs, width = 16, height = 6)
  
  # calculate AUC
  roc_calc = roc(response = dat$TrueLabel, predictor = dat$ProbabilityCTRL)
  print(roc_calc)
  
  # plot ROC curve
  auc_lab = paste0("AUC = ", round(roc_calc$auc, 4))
  roc_plot = ggplot(dat, aes(d = TrueLabel, m = ProbabilityCTRL)) +
    ggtitle("Convolutional Neural Network ROC") +
    geom_abline(aes(intercept = 0, slope = 1, color = "AUC = 0.5"), linetype = "dashed", size = 1)+
    geom_roc(aes(color = auc_lab), labels = FALSE, pointsize = 0) +
    geom_roc(linealpha = 0, n.cuts = 12, labelround = 2, labelsize = 3) +
    # scale_colour_manual(values = c("#FF9B71", "#63B0CD")) +
    scale_colour_manual(values = c("#577399", "#F39B6D")) +
    labs(x = "1 - Specificity", y = "Sensitivity", color= "Area Under the Curve (AUC)") + theme_bw() +
    theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
          axis.title.x = element_text(size=14, face="bold"),
          axis.title.y = element_text(size=14, face="bold"),
          legend.title = element_text(size=12, face="bold"),
          legend.text = element_text(size=12),
          legend.position = c(0.72, 0.14),
          legend.background = element_rect(fill = "white", color = "black"),
          panel.border = element_rect(color = "black", fill = NA, size = 1))
  
  # save ROC curve
  ggsave(file.path(wdir, "ROC Curve.pdf"), roc_plot, width = 6, height = 6)
  # write(export_interactive_roc(roc_plot), file = file.path(wdir, "ROC Curve.html"))
  
  # return data
  return(dat)
  
}

```

Create the plots specified in `plot_data` by mapping over `all`.

```{r create-plots}

# define color palette
cols = list(
  Distance = c('< 50 um' = "#F95738", '50-100 um' = "#EE964B", '> 100 um' = "#F4D35E", 'None' = "#736F72"),
  Sample = c('1190' = "#A6CEE3", '1301' = "#5D9FC9", '1619' = "#2A7FB0", '1684' = "#79B79A", '1820' = "#9ED57B", '2124' = "#5AB348", '2148' = "#619E45", '2157' = "#CC9B7F", '2169' = "#F37272", '2191' = "#E62D2F", '2207' = "#ED593B", '2242' = "#FBB268", '2250' = "#FDA13B", '2274' = "#FF7F00"),
  Condition = c(Control = "#377EB8", Alzheimer = "#CE6D8B"),
  State = c('Homeostatic' = "#39B200", 'Intermediate' = "#F0C808", 'Reactive' = "#960200")
)

# create plots
plots = imap(all, ~plot_data(.x, .y, cols))

```